Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Cancer

### 2018-05-23 submission/manuscript.tex
add references to intro

Colorectal cancer (CRC) is currently the third leading cancer-related
cause of death within the US {[}{]}. The envrionment has been attributed
as an important risk factor for CRC {[}{]}. This has lead to the
investigation of how the microbiota may exacerbate or cause tumorigensis
{[}{]} and whether the bacterial community is altered {[}{]}. Multiple
reports in case/control studies on the microbiota and CRC have
identified taxa commonly associated with the normal colon to be
decreased in patients with carcinoma tumors {[}{]}. Many of these taxa
actively produce short-chain fatty acids (SCFAs) from fiber that is part
of our general diet {[}{]}. The most extensively studied of these SCFAs
are acetate, butyrate, and propionate {[}{]}. Overall, CRC is a growing
health problem and the microbiota through SCFAs have shown promise as a
viable treatment option.

Within model systems SCFAs, in particular butyrate, have shown positive
results in CRC treatment {[}{]}. In particular, butyrate has been shown
to inhibit cancer cell growth in \emph{in vitro} systems {[}{]}.
Additionally, supplementation with food sources that bacteria use to
create these SCFAs may also be able to confer beneficial effects. Fiber
supplementation in mouse models of CRC caused an overall reduction in
tumor burden while increasing SCFA concentrations {[}{]}. Although these
model systems provide important preliminary evidence towards the ability
of SCFAs to reduce and treat tumors, the studies reporting benefit in
humans has been less convincing.

There is a lack of evidence on the benefit of increasing SCFA
concentrations to protect against CRC in human populations. The initial
case/control studies that investigated SCFA concentrations in CRC have
found that patients with carcinomas had lower concentrations of acetate,
butyrate, and propionate versus patients with adenomas and individuals
with normal colons {[}{]}. Although this would argue that increasing
SCFA concentrations could be protective against tumorigenesis, fiber
supplementation in randomized controlled trials have consistently failed
to protect against tumor recurrence {[}{]}. These findings would argue
against the utility of treatments that aim to use SCFAs to reduce or
protect against tumorigenesis. Given the lack of clear evidence in human
studies of the benefit of SCFAs in CRC there is a need for more
investigation into this area.

@@

Much of the previous literature on SCFA benefit to human colorectal
cancer has been illustrated in model systems (1). Many SCFAs are
produced through the breakdown of fiber (2). A recent study in mice
found that fiber supplementation increased SCFA concentration and
decreased tumor burden (3). Additionally, SCFAs such as butyrate have
been found to inhibit tumor growth in \emph{in vitro} experiments (4).
Yet, observations in humans has been mixed. Previous case/control
studies have suggested that SCFAs are lower in individuals with
carcinoma tumors (5). However, individual studies and meta-analysis on
fiber supplementation to prevent tumor recurrence have found no benefit
(6, 7). Our results align with what has been reported in
randomized-controlled trials that SCFAs do not protect against all types
of colorectal tumors. It is possible though that there are specific
instances where SCFAs may be beneficial.

One limitation of ours and others research into the effect of SCFAs in
colorectal cancer has been that all tumors are treated as the same type.
However, there are known differences in the types of mutations that
occur (8) and treating all tumors as equal may actually hide any benefit
that could be found in specific subsets of individuals. Similar to the
idea of specific immunotherapy for specific tumors (9), SCFAs may have
beneficial effects for specific types of colorectal cancers. Future
research will need to test if this is a valid hypothesis. Although there
are limitations with the current research on SCFAs and colorectal
tumors, evidence is building that these specific metabolites may not be
protective. Yet, taxa that are associated with SCFA production are
consistently higher in normal colons than those with carcinomas
(10--12).

The potential protection against colorectal cancers may not be from
SCFAs even though taxa associated with their production are higher in
individuals with normal colons versus patients with carcinomas (10--12).
Protection could be via a different pathway and other metabolites that
have not been extensively studied. Alternatively, protection may not
occur via a metabolite but instead occur through niche exclusion of
mouth-associated microbes (e.g. \emph{Fusobacterium},
\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (13,
14)). The idea of niche exclusion is similar to how the community
protects against \emph{Clostridium difficile} infection with chronic
inflammation replacing the role of antibiotics. Although we did not find
lower concentrations of SCFAs associated with colorectal tumors, we
think that there are many exciting new avenues to explore because of
these observations.

@@

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (15, 16). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before (pre-) and after (post-) treatment for their tumor
(n=67). For patients undergoing treatment for their tumor the length of
time between their initial and follow up sample ranged from 188 - 546
days. Our use of treatment has been previously defined as encompassing
removal of a tumor with or without chemotherapy and radiation (15).
Diagnosis of tumor was made by colonoscopic examination and
histopathological review of biopsies obtained (15, 16). The University
of Michigan Institutional Review Board approved the study and informed
consent was obtained from all participants in accordance to the
guidelines set out by the Helsinki Decleration.

\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
acetate, butyrate, and propionate followed a previously published
protocol (17). The following changes to this protocol included the use
of frozen fecal samples suspended in 1ml of PBS instead of fecal
suspensions in DNA Genotek OmniGut tubes, and the use of the acutal
weight of fecal samples instead of the average weight for SCFA
concentration normalizations. These changes did not affect the overall
median concentrations of these SCFAs between the two studies (see Table
1 (17) and Figure 1 in this report).

\textbf{16s rRNA gene sequencing.} The workflow and processing have been
described previously (15, 18, 19). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity used the OptClust algorithm (20).

\textbf{Generating imputed metagenomes.} The use of PICRUSt version
1.1.2 with the recommended standard operating protocol (21) was used.
Briefly, the mothur shared file and metadata was converted into a biom
formated table using the biom convert function, the subsequent biom file
was processed with the normalize\_by\_copy\_number.py function, and
subsequent imputed metagenomes created using the predict\_metagenomes.py
function.

\textbf{Obtaining OPFs from metagenomes.} A subset of the
cross-sectional group (n=490) containing a total of 85 individuals
(normal n=29 normal, adenoma n=28, and carcinoma n=28) was shotgun
sequenced on an Illumina HiSeq using 125 bp paired end reads and a
previously described method (22). Briefly, the sequences were quality
filtered and sequences aligning to the human genome were removed prior
to contig assembly with MEGAHIT (23). Open Reading Frames (ORFs) were
identified using Prodigal (24), counts generated using Diamond (25),
subsequent clustering into Operational Protein Families (OPFs) used
mmseq2 (26), and OPF alignment used the KEGG database (27).

\textbf{Pulling genes involved with SCFA synthesis.} Specific genes
located near the end of the pathways involved in the synthesis of
acetate, butyrate, and propionate were analyzed for any differences
between individuals with normal colons and those with tumors. These
genes were based on pathways from KEGG as well as previous research (27,
28) and a list can be found in the supplemental material {[}Table S1{]}.

\textbf{Random Forest Models.} The model was first trained on 80\% of
the data and then tested on the held out 20\% (80/20 split) using the
Random Forest algorithm for both classification and regression models
(29). This was repeated on 100 differen 80/20 splits of the data to
generate a reasonable range for the AUC of the model. The reported AUCs,
unless otherwise specified, are for the test sets. Classification models
were built to group normal versus adenoma and normal versus carcinoma or
high versus low SCFA concentrations.

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (30). Generally, differences between the
different disease groups used a Kruskal-Walis rank sum test with a
Dunn's post-hoc test. Where appropriate Benjamini-Hochberg was used to
correct for multiple comparisons (31). First, we assessed differences in
SCFA concentrations measured by HPLC between individuals with normal
colons and patients with tumors (adenoma or carcinoma). We then analyzed
whether SCFA concentrations changed in patients with an adenoma or
carcinoma pre- versus post-treatment. Next, we assessed whether OTUs
alone or OTUs and SCFAs were better able to classify individuals with
and without tumor using Random Forest models. Next, the imputed gene
counts of important mediators of SCFA creation were tested. Finally, the
counts generated for OPFs that matched important genes involved with
SCFA creation were analyzed. Finally, models to classify high or low
SCFA concentration based on 16S rRNA gene sequencing data were created
using the Random Forest algorithm.

@@

\hypertarget{refs}{}
\hypertarget{ref-test_OKeefe2016}{}
1. \textbf{O'Keefe SJD}. 2016. Diet, microorganisms and their
metabolites and colon cancer. Nature Reviews Gastroenterology \&
Hepatology \textbf{13}:691--706.
doi:\href{https://doi.org/10.1038/nrgastro.2016.165}{10.1038/nrgastro.2016.165}.

\hypertarget{ref-Holscher2017}{}
2. \textbf{Holscher HD}. 2017. Dietary fiber and prebiotics and the
gastrointestinal microbiota. Gut Microbes \textbf{8}:172--184.
doi:\href{https://doi.org/10.1080/19490976.2017.1290756}{10.1080/19490976.2017.1290756}.

\hypertarget{ref-Bishehsari2018}{}
3. \textbf{Bishehsari F}, \textbf{Engen P}, \textbf{Preite N},
\textbf{Tuncil Y}, \textbf{Naqib A}, \textbf{Shaikh M}, \textbf{Rossi
M}, \textbf{Wilber S}, \textbf{Green S}, \textbf{Hamaker B},
\textbf{Khazaie K}, \textbf{Voigt R}, \textbf{Forsyth C},
\textbf{Keshavarzian A}. 2018. Dietary fiber treatment corrects the
composition of gut microbiota, promotes SCFA production, and suppresses
colon carcinogenesis. Genes \textbf{9}:102.
doi:\href{https://doi.org/10.3390/genes9020102}{10.3390/genes9020102}.

\hypertarget{ref-Encarnao2018}{}
4. \textbf{Encarnação JC}, \textbf{Pires AS}, \textbf{Amaral RA},
\textbf{Gonçalves TJ}, \textbf{Laranjo M}, \textbf{Casalta-Lopes JE},
\textbf{Gonçalves AC}, \textbf{Sarmento-Ribeiro AB}, \textbf{Abrantes
AM}, \textbf{Botelho MF}. 2018. Butyrate, a dietary fiber derivative
that improves irinotecan effect in colon cancer cells. The Journal of
Nutritional Biochemistry \textbf{56}:183--192.
doi:\href{https://doi.org/10.1016/j.jnutbio.2018.02.018}{10.1016/j.jnutbio.2018.02.018}.

\hypertarget{ref-Ohigashi2013}{}
5. \textbf{Ohigashi S}, \textbf{Sudo K}, \textbf{Kobayashi D},
\textbf{Takahashi O}, \textbf{Takahashi T}, \textbf{Asahara T},
\textbf{Nomoto K}, \textbf{Onodera H}. 2013. Changes of the intestinal
microbiota, short chain fatty acids, and fecal pH in patients with
colorectal cancer. Digestive Diseases and Sciences
\textbf{58}:1717--1726.
doi:\href{https://doi.org/10.1007/s10620-012-2526-4}{10.1007/s10620-012-2526-4}.

\hypertarget{ref-Schatzkin2000}{}
6. \textbf{Schatzkin A}, \textbf{Lanza E}, \textbf{Corle D},
\textbf{Lance P}, \textbf{Iber F}, \textbf{Caan B}, \textbf{Shike M},
\textbf{Weissfeld J}, \textbf{Burt R}, \textbf{Cooper MR},
\textbf{Kikendall JW}, \textbf{Cahill J}, \textbf{Freedman L},
\textbf{Marshall J}, \textbf{Schoen RE}, \textbf{Slattery M}. 2000. Lack
of effect of a low-fat, high-fiber diet on the recurrence of colorectal
adenomas. New England Journal of Medicine \textbf{342}:1149--1155.
doi:\href{https://doi.org/10.1056/nejm200004203421601}{10.1056/nejm200004203421601}.

\hypertarget{ref-Yao2017}{}
7. \textbf{Yao Y}, \textbf{Suo T}, \textbf{Andersson R}, \textbf{Cao Y},
\textbf{Wang C}, \textbf{Lu J}, \textbf{Chui E}. 2017. Dietary fibre for
the prevention of recurrent colorectal adenomas and carcinomas. Cochrane
Database of Systematic Reviews.
doi:\href{https://doi.org/10.1002/14651858.cd003430.pub2}{10.1002/14651858.cd003430.pub2}.

\hypertarget{ref-Fearon1990}{}
8. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
colorectal tumorigenesis. Cell \textbf{61}:759--767.
doi:\href{https://doi.org/10.1016/0092-8674(90)90186-i}{10.1016/0092-8674(90)90186-i}.

\hypertarget{ref-Thomas2016}{}
9. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
agents targeting the BCR-ABL tyrosine kinase as philadelphia
chromosome-positive acute lymphoblastic leukemia treatment. Expert
Opinion on Drug Discovery \textbf{11}:1061--1070.
doi:\href{https://doi.org/10.1080/17460441.2016.1227318}{10.1080/17460441.2016.1227318}.

\hypertarget{ref-Shah2017}{}
10. \textbf{Shah MS}, \textbf{DeSantis TZ}, \textbf{Weinmaier T},
\textbf{McMurdie PJ}, \textbf{Cope JL}, \textbf{Altrichter A},
\textbf{Yamal J-M}, \textbf{Hollister EB}. 2017. Leveraging
sequence-based faecal microbial community survey data to identify a
composite biomarker for colorectal cancer. Gut \textbf{67}:882--891.
doi:\href{https://doi.org/10.1136/gutjnl-2016-313189}{10.1136/gutjnl-2016-313189}.

\hypertarget{ref-normalization_Sze2017}{}
11. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
microbiota in patients after treatment for colonic lesions. Microbiome
\textbf{5}.
doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.

\hypertarget{ref-meta_analysis_crc_Sze2018}{}
12. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
rRNA gene surveys to identify reproducible biomarkers in individuals
with colorectal tumors.
doi:\href{https://doi.org/10.1101/285486}{10.1101/285486}.

\hypertarget{ref-Zeller2014}{}
13. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
\textbf{Amiot A}, \textbf{Bohm J}, \textbf{Brunetti F},
\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Nhieu JTV},
\textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
for early-stage detection of colorectal cancer. Molecular Systems
Biology \textbf{10}:766--766.
doi:\href{https://doi.org/10.15252/msb.20145645}{10.15252/msb.20145645}.

\hypertarget{ref-Baxter2016}{}
14. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\hypertarget{ref-normalization_sze2017}{}
15. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
microbiota in patients after treatment for colonic lesions. Microbiome
\textbf{5}.
doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.

\hypertarget{ref-crc_model_baxter2016}{}
16. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\hypertarget{ref-scfa_measures_venkataraman2016}{}
17. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
\textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
Variable responses of human microbiomes to dietary supplementation with
resistant starch. Microbiome \textbf{4}.
doi:\href{https://doi.org/10.1186/s40168-016-0178-x}{10.1186/s40168-016-0178-x}.

>>>>

Colorectal cancer (CRC) is currently the third leading cancer-related
cause of death within the US (1, 2). The envrionment has been attributed
as an important risk factor for CRC (3). This has lead to the
investigation of how the microbiota may exacerbate or cause tumorigensis
(4--6) and whether the bacterial community is altered (7, 8). Multiple
reports in case/control studies on the microbiota and CRC have
identified taxa commonly associated with the normal colon to be
decreased in patients with carcinoma tumors (9--11). Many of these taxa
actively produce short-chain fatty acids (SCFAs) from fiber that is part
of our general diet (12). The most extensively studied of these SCFAs
are acetate, butyrate, and propionate (13). Overall, CRC is a growing
health problem and the microbiota through SCFAs have shown promise as a
viable treatment option.

Within model systems SCFAs, in particular butyrate, have shown positive
results in CRC treatment (14). In particular, butyrate has been shown to
inhibit cancer cell growth in \emph{in vitro} systems (15).
Additionally, supplementation with food sources that bacteria use to
create these SCFAs may also be able to confer beneficial effects. Fiber
supplementation in mouse models of CRC caused an overall reduction in
tumor burden while also increasing SCFA concentrations (16). Although
these model systems provide important preliminary evidence towards the
ability of SCFAs to reduce and treat tumors, the studies reporting
benefit in humans has been less convincing.

There is a lack of evidence on the benefit of increasing SCFA
concentrations to protect against CRC in human populations. The initial
case/control studies that investigated SCFA concentrations in CRC have
found that patients with carcinomas had lower concentrations of acetate,
butyrate, and propionate versus patients with adenomas and individuals
with normal colons (17). Although this would argue that increasing SCFA
concentrations could be protective against tumorigenesis, fiber
supplementation in randomized controlled trials have consistently failed
to protect against tumor recurrence (18). These findings would argue
against the utility of treatments that aim to use SCFAs to reduce or
protect against tumorigenesis. Given the lack of clear evidence in human
studies of the benefit of SCFAs in CRC there is a need for more
investigation into this area.

@@

Much of the previous literature on SCFA benefit to human colorectal
cancer has been illustrated in model systems (14). Many SCFAs are
produced through the breakdown of fiber (12). A recent study in mice
found that fiber supplementation increased SCFA concentration and
decreased tumor burden (16). Additionally, SCFAs such as butyrate have
been found to inhibit tumor growth in \emph{in vitro} experiments (15).
Yet, observations in humans has been mixed. Previous case/control
studies have suggested that SCFAs are lower in individuals with
carcinoma tumors (17). However, individual studies and meta-analysis on
fiber supplementation to prevent tumor recurrence have found no benefit
(18, 19). Our results align with what has been reported in
randomized-controlled trials that SCFAs do not protect against all types
of colorectal tumors. It is possible though that there are specific
instances where SCFAs may be beneficial.

One limitation of ours and others research into the effect of SCFAs in
colorectal cancer has been that all tumors are treated as the same type.
However, there are known differences in the types of mutations that
occur (20) and treating all tumors as equal may actually hide any
benefit that could be found in specific subsets of individuals. Similar
to the idea of specific immunotherapy for specific tumors (21), SCFAs
may have beneficial effects for specific types of colorectal cancers.
Future research will need to test if this is a valid hypothesis.
Although there are limitations with the current research on SCFAs and
colorectal tumors, evidence is building that these specific metabolites
may not be protective. Yet, taxa that are associated with SCFA
production are consistently higher in normal colons than those with
carcinomas (7, 8, 22).

The potential protection against colorectal cancers may not be from
SCFAs even though taxa associated with their production are higher in
individuals with normal colons versus patients with carcinomas (7, 8,
22). Protection could be via a different pathway and other metabolites
that have not been extensively studied. Alternatively, protection may
not occur via a metabolite but instead occur through niche exclusion of
mouth-associated microbes (e.g. \emph{Fusobacterium},
\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (9,
10)). The idea of niche exclusion is similar to how the community
protects against \emph{Clostridium difficile} infection with chronic
inflammation replacing the role of antibiotics. Although we did not find
lower concentrations of SCFAs associated with colorectal tumors, we
think that there are many exciting new avenues to explore because of
these observations.

@@

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (23, 24). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before (pre-) and after (post-) treatment for their tumor
(n=67). For patients undergoing treatment for their tumor the length of
time between their initial and follow up sample ranged from 188 - 546
days. Our use of treatment has been previously defined as encompassing
removal of a tumor with or without chemotherapy and radiation (23).
Diagnosis of tumor was made by colonoscopic examination and
histopathological review of biopsies obtained (23, 24). The University
of Michigan Institutional Review Board approved the study and informed
consent was obtained from all participants in accordance to the
guidelines set out by the Helsinki Decleration.

\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
acetate, butyrate, and propionate followed a previously published
protocol (25). The following changes to this protocol included the use
of frozen fecal samples suspended in 1ml of PBS instead of fecal
suspensions in DNA Genotek OmniGut tubes, and the use of the acutal
weight of fecal samples instead of the average weight for SCFA
concentration normalizations. These changes did not affect the overall
median concentrations of these SCFAs between the two studies (see Table
1 (25) and Figure 1 in this report).

\textbf{16s rRNA gene sequencing.} The workflow and processing have been
described previously (23, 26, 27). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity used the OptClust algorithm (28).

\textbf{Generating imputed metagenomes.} The use of PICRUSt version
1.1.2 with the recommended standard operating protocol (29) was used.
Briefly, the mothur shared file and metadata was converted into a biom
formated table using the biom convert function, the subsequent biom file
was processed with the normalize\_by\_copy\_number.py function, and
subsequent imputed metagenomes created using the predict\_metagenomes.py
function.

\textbf{Obtaining OPFs from metagenomes.} A subset of the
cross-sectional group (n=490) containing a total of 85 individuals
(normal n=29 normal, adenoma n=28, and carcinoma n=28) was shotgun
sequenced on an Illumina HiSeq using 125 bp paired end reads and a
previously described method (30). Briefly, the sequences were quality
filtered and sequences aligning to the human genome were removed prior
to contig assembly with MEGAHIT (31). Open Reading Frames (ORFs) were
identified using Prodigal (32), counts generated using Diamond (33),
subsequent clustering into Operational Protein Families (OPFs) used
mmseq2 (34), and OPF alignment used the KEGG database (35).

\textbf{Pulling genes involved with SCFA synthesis.} Specific genes
located near the end of the pathways involved in the synthesis of
acetate, butyrate, and propionate were analyzed for any differences
between individuals with normal colons and those with tumors. These
genes were based on pathways from KEGG as well as previous research (35,
36) and a list can be found in the supplemental material {[}Table S1{]}.

\textbf{Random Forest Models.} The model was first trained on 80\% of
the data and then tested on the held out 20\% (80/20 split) using the
Random Forest algorithm for both classification and regression models
(37). This was repeated on 100 differen 80/20 splits of the data to
generate a reasonable range for the AUC of the model. The reported AUCs,
unless otherwise specified, are for the test sets. Classification models
were built to group normal versus adenoma and normal versus carcinoma or
high versus low SCFA concentrations.

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (38). Generally, differences between the
different disease groups used a Kruskal-Walis rank sum test with a
Dunn's post-hoc test. Where appropriate Benjamini-Hochberg was used to
correct for multiple comparisons (39). First, we assessed differences in
SCFA concentrations measured by HPLC between individuals with normal
colons and patients with tumors (adenoma or carcinoma). We then analyzed
whether SCFA concentrations changed in patients with an adenoma or
carcinoma pre- versus post-treatment. Next, we assessed whether OTUs
alone or OTUs and SCFAs were better able to classify individuals with
and without tumor using Random Forest models. Next, the imputed gene
counts of important mediators of SCFA creation were tested. Finally, the
counts generated for OPFs that matched important genes involved with
SCFA creation were analyzed. Finally, models to classify high or low
SCFA concentration based on 16S rRNA gene sequencing data were created
using the Random Forest algorithm.

@@

\hypertarget{refs}{}
\hypertarget{ref-Haggar2009}{}
1. \textbf{Haggar F}, \textbf{Boushey R}. 2009. Colorectal cancer
epidemiology: Incidence, mortality, survival, and risk factors. Clinics
in Colon and Rectal Surgery \textbf{22}:191--197.
doi:\href{https://doi.org/10.1055/s-0029-1242458}{10.1055/s-0029-1242458}.

\hypertarget{ref-Siegel2016}{}
2. \textbf{Siegel RL}, \textbf{Miller KD}, \textbf{Jemal A}. 2016.
Cancer statistics, 2016. CA: A Cancer Journal for Clinicians
\textbf{66}:7--30.
doi:\href{https://doi.org/10.3322/caac.21332}{10.3322/caac.21332}.

\hypertarget{ref-Lichtenstein2000}{}
3. \textbf{Lichtenstein P}, \textbf{Holm NV}, \textbf{Verkasalo PK},
\textbf{Iliadou A}, \textbf{Kaprio J}, \textbf{Koskenvuo M},
\textbf{Pukkala E}, \textbf{Skytthe A}, \textbf{Hemminki K}. 2000.
Environmental and heritable factors in the causation of cancer analyses
of cohorts of twins from sweden, denmark, and finland. New England
Journal of Medicine \textbf{343}:78--85.
doi:\href{https://doi.org/10.1056/nejm200007133430201}{10.1056/nejm200007133430201}.

\hypertarget{ref-Zackular2013}{}
4. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Iverson KD},
\textbf{Sadler WD}, \textbf{Petrosino JF}, \textbf{Chen GY},
\textbf{Schloss PD}. 2013. The gut microbiome modulates colon
tumorigenesis. mBio \textbf{4}:e00692--13--e00692--13.
doi:\href{https://doi.org/10.1128/mbio.00692-13}{10.1128/mbio.00692-13}.

\hypertarget{ref-Baxter2014}{}
5. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
\textbf{Schloss PD}. 2014. Structure of the gut microbiome following
colonization with human feces determines colonic tumor burden.
Microbiome \textbf{2}:20.
doi:\href{https://doi.org/10.1186/2049-2618-2-20}{10.1186/2049-2618-2-20}.

\hypertarget{ref-Zackular2015}{}
6. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Chen GY},
\textbf{Schloss PD}. 2015. Manipulation of the gut microbiota reveals
role in colon tumorigenesis. mSphere \textbf{1}:e00001--15.
doi:\href{https://doi.org/10.1128/msphere.00001-15}{10.1128/msphere.00001-15}.

\hypertarget{ref-Shah2017}{}
7. \textbf{Shah MS}, \textbf{DeSantis TZ}, \textbf{Weinmaier T},
\textbf{McMurdie PJ}, \textbf{Cope JL}, \textbf{Altrichter A},
\textbf{Yamal J-M}, \textbf{Hollister EB}. 2017. Leveraging
sequence-based faecal microbial community survey data to identify a
composite biomarker for colorectal cancer. Gut \textbf{67}:882--891.
doi:\href{https://doi.org/10.1136/gutjnl-2016-313189}{10.1136/gutjnl-2016-313189}.

\hypertarget{ref-meta_analysis_crc_Sze2018}{}
8. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
rRNA gene surveys to identify reproducible biomarkers in individuals
with colorectal tumors.
doi:\href{https://doi.org/10.1101/285486}{10.1101/285486}.

\hypertarget{ref-Zeller2014}{}
9. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
\textbf{Amiot A}, \textbf{Bohm J}, \textbf{Brunetti F},
\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Nhieu JTV},
\textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
for early-stage detection of colorectal cancer. Molecular Systems
Biology \textbf{10}:766--766.
doi:\href{https://doi.org/10.15252/msb.20145645}{10.15252/msb.20145645}.

\hypertarget{ref-Baxter2016}{}
10. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\hypertarget{ref-Kostic2011}{}
11. \textbf{Kostic AD}, \textbf{Gevers D}, \textbf{Pedamallu CS},
\textbf{Michaud M}, \textbf{Duke F}, \textbf{Earl AM}, \textbf{Ojesina
AI}, \textbf{Jung J}, \textbf{Bass AJ}, \textbf{Tabernero J},
\textbf{Baselga J}, \textbf{Liu C}, \textbf{Shivdasani RA},
\textbf{Ogino S}, \textbf{Birren BW}, \textbf{Huttenhower C},
\textbf{Garrett WS}, \textbf{Meyerson M}. 2011. Genomic analysis
identifies association of fusobacterium with colorectal carcinoma.
Genome Research \textbf{22}:292--298.
doi:\href{https://doi.org/10.1101/gr.126573.111}{10.1101/gr.126573.111}.

\hypertarget{ref-Holscher2017}{}
12. \textbf{Holscher HD}. 2017. Dietary fiber and prebiotics and the
gastrointestinal microbiota. Gut Microbes \textbf{8}:172--184.
doi:\href{https://doi.org/10.1080/19490976.2017.1290756}{10.1080/19490976.2017.1290756}.

\hypertarget{ref-Louis2016}{}
13. \textbf{Louis P}, \textbf{Flint HJ}. 2016. Formation of propionate
and butyrate by the human colonic microbiota. Environmental Microbiology
\textbf{19}:29--41.
doi:\href{https://doi.org/10.1111/1462-2920.13589}{10.1111/1462-2920.13589}.

\hypertarget{ref-test_OKeefe2016}{}
14. \textbf{O'Keefe SJD}. 2016. Diet, microorganisms and their
metabolites and colon cancer. Nature Reviews Gastroenterology \&
Hepatology \textbf{13}:691--706.
doi:\href{https://doi.org/10.1038/nrgastro.2016.165}{10.1038/nrgastro.2016.165}.

\hypertarget{ref-Encarnao2018}{}
15. \textbf{Encarnação JC}, \textbf{Pires AS}, \textbf{Amaral RA},
\textbf{Gonçalves TJ}, \textbf{Laranjo M}, \textbf{Casalta-Lopes JE},
\textbf{Gonçalves AC}, \textbf{Sarmento-Ribeiro AB}, \textbf{Abrantes
AM}, \textbf{Botelho MF}. 2018. Butyrate, a dietary fiber derivative
that improves irinotecan effect in colon cancer cells. The Journal of
Nutritional Biochemistry \textbf{56}:183--192.
doi:\href{https://doi.org/10.1016/j.jnutbio.2018.02.018}{10.1016/j.jnutbio.2018.02.018}.

\hypertarget{ref-Bishehsari2018}{}
16. \textbf{Bishehsari F}, \textbf{Engen P}, \textbf{Preite N},
\textbf{Tuncil Y}, \textbf{Naqib A}, \textbf{Shaikh M}, \textbf{Rossi
M}, \textbf{Wilber S}, \textbf{Green S}, \textbf{Hamaker B},
\textbf{Khazaie K}, \textbf{Voigt R}, \textbf{Forsyth C},
\textbf{Keshavarzian A}. 2018. Dietary fiber treatment corrects the
composition of gut microbiota, promotes SCFA production, and suppresses
colon carcinogenesis. Genes \textbf{9}:102.
doi:\href{https://doi.org/10.3390/genes9020102}{10.3390/genes9020102}.

\hypertarget{ref-Ohigashi2013}{}
17. \textbf{Ohigashi S}, \textbf{Sudo K}, \textbf{Kobayashi D},
\textbf{Takahashi O}, \textbf{Takahashi T}, \textbf{Asahara T},
\textbf{Nomoto K}, \textbf{Onodera H}. 2013. Changes of the intestinal
microbiota, short chain fatty acids, and fecal pH in patients with
colorectal cancer. Digestive Diseases and Sciences
\textbf{58}:1717--1726.
doi:\href{https://doi.org/10.1007/s10620-012-2526-4}{10.1007/s10620-012-2526-4}.

\hypertarget{ref-Yao2017}{}
18. \textbf{Yao Y}, \textbf{Suo T}, \textbf{Andersson R}, \textbf{Cao
Y}, \textbf{Wang C}, \textbf{Lu J}, \textbf{Chui E}. 2017. Dietary fibre
for the prevention of recurrent colorectal adenomas and carcinomas.
Cochrane Database of Systematic Reviews.
doi:\href{https://doi.org/10.1002/14651858.cd003430.pub2}{10.1002/14651858.cd003430.pub2}.

\hypertarget{ref-Schatzkin2000}{}
19. \textbf{Schatzkin A}, \textbf{Lanza E}, \textbf{Corle D},
\textbf{Lance P}, \textbf{Iber F}, \textbf{Caan B}, \textbf{Shike M},
\textbf{Weissfeld J}, \textbf{Burt R}, \textbf{Cooper MR},
\textbf{Kikendall JW}, \textbf{Cahill J}, \textbf{Freedman L},
\textbf{Marshall J}, \textbf{Schoen RE}, \textbf{Slattery M}. 2000. Lack
of effect of a low-fat, high-fiber diet on the recurrence of colorectal
adenomas. New England Journal of Medicine \textbf{342}:1149--1155.
doi:\href{https://doi.org/10.1056/nejm200004203421601}{10.1056/nejm200004203421601}.
